### 2022-04-17 doc/university-covid.tex
bunch of edits

Mathematical models have played a crucial role in guiding pandemic responses; university campuses provide a particularly well documented case for institutional outbreaks.
Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions for the Princeton University campus.
We demonstrate strong spatiotemporal correlations in epidemic patterns between Princeton University and surrounding communities.
These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic.
Mathematical modeling further illustrates that an increase in the number of cases in November 2021 is consistent with waning immunity.
Predicting future outbreaks, and the effectiveness of intervention measures on campus, is then expected to be sensitive to underlying epidemiological conditions, including community prevalence and immunity landscape.

@@

University campuses also present unique challenges to controlling an outbreak.
High-density environments, especially within dorms, and a large fraction of asymptomatic infections (due to the young age of university students) can readily permit rapid transmission.
These kinds of contacts are inherently difficult to trace, making contact tracing less effective.
The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period despite similar levels of testing and vaccination.
Here, we choose to focus on the spread of SARS-CoV-2 on PU campus alone to eliminate heterogeneities inherent to such comparisons.

In this study, we analyze the spread of SARS-CoV-2 on PU campus (\fref{princeton}).
We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
PU is located in Mercer County, New Jersey, USA;
the population comprises of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
For simplicity, we divide the epidemic into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
Throughout the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Asymptomatic individuals who tested positive were required to isolate for 10 days.
Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
PCR positives were exempt from asymptomatic testing for 90 days.
Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University (PU).}
(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
Red lines represent the weekly number of cases in Mercer County.
Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
(D--F) Correlations between the weekly number of cases in PU and in Mercer County.
Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
(G) Ratios between weekly numbers of cases reported from PU and Mercer County.
Solid lines represent the observed ratios.
Dotted lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
\label{fig:princeton}
}
\end{figure}

During fall 2020, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
Undergraduate students were not allowed to return to campus as all classes were held virtually, and so their numbers were minimal on campus ($<300$).  
Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
A sudden decrease in the number of cases around Thanksgiving partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).

In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).

In fall 2021, all students and faculty and staff members were required to be vaccinated \swp{with two doses? need to ask Robin} with a very few medical and religious exemptions.
By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were vaccinated \swp{ask Robin}.
As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of faculty and staff members were vaccinated.
Vaccinees were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
In-person classes and social events fully resumed on campus while all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
The number of cases remained similar to previous semesters until November when the number of cases began to increase primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
In order to prevent transmission, testing frequency was increased to twice a week for undergraduate students on November 27th, 2021; the size of non-academic gatherings were also limited to 20 people.
The number of cases decreased slightly as classes ended but soon increased back up as the Omicron variant began to spread on campus and in Mercer county.

Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from PU and those from Mercer county:
fall 2020 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}D); spring 2020 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}E); and fall 2021 ($\rho = 0.90$ (95\% CI: 0.78--0.96; $p < 0.001$), \fref{princeton}F). 
These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).

These correlations likely reflect commuting and contact patterns, and therefore we expect SARS-CoV-2 dynamics on campus to be correlated with those from nearby large cities as well. 
We find similarly strong correlations with New York City: fall 2020 ($\rho = 0.74$ (95\% CI: 0.44--0.90; $p < 0.001$), Supplementary Figure S3A); spring 2020 ($\rho = 0.84$ (95\% CI: 0.61--0.94; $p < 0.001$), Supplementary Figure S3B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.70--0.94; $p < 0.001$), Supplementary Figure S3C).
The same picture emerges for Philadelphia except for spring 2020: fall 2020 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.68--0.94; $p < 0.001$), Supplementary Figure S4C).
Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S5):
epidemic dynamics were highly synchronized across all counties in fall 2020 and became less synchronized over time. 
These correlations significantly decreased with distance in spring 2020 ($\rho = -0.36$ (95\% CI: $-0.49$--$-0.21$; $p < 0.001$)) and fall 20201 ($\rho = -0.58$ (95\% CI: $-0.68$--$-0.46$; $p < 0.001$)).
These variations likely reflect differences in vaccination levels and the timing of the introduction of the Omicron variant.

Finally, mass testing allows us to infer the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission provided that testing patterns remain constant in both places.
We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in PU during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
We also find that some variations in the ratio around the mean can be explained by holiday effects.
For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.

@@

We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 on the PU campus.
This model was initially developed and used throughout the pandemic to inform policy decisions in PU, including the frequency of asymptomatic tests and the number of isolation beds required.
We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus after a virtual semester) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in PU without an over-proliferation of parameters.
The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
Previous versions of the model included contact tracing, but we exclude it in this model for simplicity.

Infection processes are modeled based on standard compartmental structures (Supplementary Figure S6).
Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
Exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections for $D_p = 3$ days on average.
Presymptomatic individuals can then either remain asymptomatic with probability $p_a = 0.4$ or develop symptoms with the remaining probability of $1-p_a = 0.6$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s=3$) and equal transmission rates.
Recovered individuals are assumed to be immune to reinfections throughout a semester.
Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}:
more specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
Assumed parameters are broadly consistent with other modeling approaches  \citep{brett2020transmission,lavezzo2020suppression}.

Transmission processes are modeled by first setting the contact reproduction number $\Rc$, which we define as the average number of infectious contacts an infected individual would make throughout the course of their infection;
here, infectious contacts refer to contacts that would result in infection when the contactee is susceptible to infection.
The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
We further decompose $\Rc$ into pre-symptomatic $\mathcal R_p = \beta_p D_p$ and (a)symptomatic $\mathcal R_s = \beta_s D_s$ reproduction numbers, where $\beta_p$ and $\beta_s$ represent the corresponding infectious contact rates during pre-symptomatic and (a)symptomatic stages, respectively. Pre-symptomatic and (a)symptomatic reproduction numbers are calculated based on the assumed value of the proportion of presymptomatic transmission $p_p = 0.5$: $\mathcal R_p/\mathcal R_s = p_p/(1 - p_p)$.
On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events \citep{10.12688/wellcomeopenres.15842.3}.
Infectious contacts from the community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, contactees have a reduced probability of infection corresponding to their susceptibility (discussed later).

@@

In this study, we use this model to retrospectively analyze past outbreaks and further predict future epidemic trajectories for the Omicron variant.
First, we try to match our model to epidemic patterns seen on campus by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume that all other parameters are equal across different groups for simplicity. 
We further assume that the population mixes homogeneously.
While these assumptions are most conservative, epidemiological parameters  and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.

@@

\section*{Future projections of the Omicron variant}

>>>>

Mathematical models have played a crucial role in guiding pandemic responses; university campuses provide a particularly well documented case for institutional outbreaks.
Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions for the Princeton University campus.
We demonstrate strong spatiotemporal correlations in epidemic patterns between Princeton University and surrounding communities.
These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic.

@@

University campuses present unique challenges to controlling an outbreak.
High-density environments, especially within dorms, and a large fraction of asymptomatic infections (due to the young age of university students) can readily permit rapid transmission.
These kinds of contacts are inherently difficult to keep track of, making contact tracing less effective.
The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly asymptomatic testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period with similar levels of testing and vaccination.
Here, we focused on the spread of SARS-CoV-2 on PU campus alone to eliminate heterogeneities inherent to such comparisons.

In this study, we analyzed the spread of SARS-CoV-2 on PU campus (\fref{princeton}).
We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
PU is located in Mercer County, New Jersey, USA;
the population comprises of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
Throughout the majority of the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
Asymptomatic individuals who tested positive were required to isolate for 10 days.
Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{Robin and Irini: is self-isolate a better term?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
PCR positives were exempt from asymptomatic testing for 90 days.
Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine statuses are up-to-date, which in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University (PU).}
(A--D) Epidemic trajectories across four semesters: Fall 2020--2021 (A), Spring 2020--2021 (B), Fall 2021--2022 (C), and Spring 2021--2022 (D).
Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
Red lines represent the weekly number of cases in Mercer County.
Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
(E--H) Correlations between the weekly number of cases in PU and in Mercer County.
Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
(I--L) Ratios between weekly numbers of cases reported from PU and Mercer County.
The shaded orange area covers the region between 1:100 and 1:25 ratios.
\label{fig:princeton}
}
\end{figure}

During the fall semester of the 2020--2021 academic year, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
Undergraduate students were not allowed to return to campus as all classes were held virtually, and so their numbers were minimal on campus ($<300$).  
Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
A sudden decrease in the number of cases around Thanksgiving partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).

In the beginning of the spring semester of the 2020--2021 academic year, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
Returning students were required to be tested and quarantine \swp{Robin and Irini: for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).

For the fall semester of the 2021--2022 academic year, all students and faculty and staff members were required to be vaccinated \swp{with two doses? need to ask Robin and Irini} with a very few medical and religious exemptions.
By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were vaccinated \swp{ask Robin and Irini}.
% As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of faculty and staff members were vaccinated.
Vaccinees were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
In-person classes and social events fully resumed on campus while all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
The number of cases remained similar to previous semesters until November when the number of cases began to increase primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
In order to prevent transmission, testing frequency was increased to twice a week for undergraduate students on November 27th, 2021; the size of non-academic gatherings were also limited to 20 people.
The number of cases decreased slightly as classes ended but soon increased back up as the Omicron (BA.1) variant began to spread on campus and in Mercer county.

For the spring semester of the 2021--2022 academic year, all eligible students and faculty and staff members were required to be boosted.
By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were boosted \swp{ask Robin and Irini}.
Undergraduate students were still required to be tested twice a week to prevent the additional spread of the Omicron variant.
The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
As the campus and local cases numbers decreased, the gathering policy was updated to allow food in events were no longer limited to 20 people;
in addition, the testing frequency was reduced a once a week.
Following the policy change, a large outbreak occurred on campus with high case numbers persisting until the spring break.
On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
The timing of this outbreak also coincided with a rapid turnover of the Omicron subvariant BA.2---the proportion of BA.2 subvariant reached ??\% compared to ??\% from the previous week \swp{Ask Dan}.

Across the first three semesters, we find strong and significant correlations between the weekly logged numbers of cases from PU and those from Mercer county:
fall 2020--2021 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.86$ (95\% CI: 0.67--0.94; $p < 0.001$), \fref{princeton}G). 
These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.32$ (95\% CI: -0.34--0.77; $p = 0.33$).
Stratifying cases by the population shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S1). 
We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
Across the first three semesters, both the total cases and faculty and staff cases showed similar levels of correlations with local cases. 
For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other countries ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S2).

These correlations likely reflect commuting and contact patterns, and therefore we expect SARS-CoV-2 dynamics on campus to be correlated with those from nearby large cities as well. 
We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.74$ (95\% CI: 0.44--0.90; $p < 0.001$), Supplementary Figure S3A); spring 2020--2021 ($\rho = 0.84$ (95\% CI: 0.61--0.94; $p < 0.001$), Supplementary Figure S3B); fall 2021--2022 ($\rho = 0.78$ (95\% CI: 0.51--0.91; $p < 0.001$), Supplementary Figure S3C); and spring 2021--2022 ($\rho = 0.33$ (95\% CI: $-0.34$--0.78; $p=0.3$), Supplementary Figure S3D).

The same picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020--2021 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); fall 2021--2022 ($\rho = 0.78$ (95\% CI: 0.51--0.91; $p < 0.001$), Supplementary Figure S4C); 
and spring 2021--2022 ($\rho=0.37$ (95\% CI: $-0.30$--0.80; $p=0.27$), Supplementary Figure S4D).
Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S5):
epidemic dynamics were highly synchronized across all counties in fall 2020--2021 and became less synchronized over time. 
These correlations significantly decreased with distance in spring 2020--2021 ($\rho = -0.36$ (95\% CI: $-0.49$--$-0.21$; $p < 0.001$)) and fall 2021--2022 ($\rho = -0.58$ (95\% CI: $-0.68$--$-0.46$; $p < 0.001$)).
These variations likely reflect differences in vaccination levels and the timing of the introduction of the Omicron variant.

Finally, mass testing allows us to infer the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission provided that testing patterns remain constant in both places.
We find that the ratio between the weekly numbers of cases from Princeton and those from Mercer county stayed between 1:100 and 1:25 until the end of the fall semester of the 2021--2022 academic year (\fref{princeton}I--K).
An increase in this ratio at the end of November 2021 was associated with the campus outbreak before Thanksgiving followed by an introduction of the Omicron variant in December---this deviation indicates an increase in the amount of transmission on campus.
During the spring semester of the 2021--2022 academic year, the ratio between PU cases and Mercer county cases increased above one, meaning that more cases were reported in PU than elsewhere in Mercer county (\fref{princeton}L);
notably, we did not see an increase in Mercer county cases, meaning that there was little-to-no transmission from campus to local community.

% We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
% Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
% However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in PU during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
% We also find that some variations in the ratio around the mean can be explained by holiday effects.
% For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
% Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.

@@

We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 on the PU campus.
This model was initially developed and used throughout the pandemic to inform policy decisions in PU, including the frequency of asymptomatic tests and the number of isolation beds required.
We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus after a virtual semester) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in PU without an over-proliferation of parameters.
The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
Previous versions of the model included contact tracing, but we exclude it in this study for simplicity.

Infection processes are modeled based on standard compartmental structures (Supplementary Figure S6).
Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
Exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections for $D_p = 3$ days on average.
Presymptomatic individuals can then either remain asymptomatic with probability $p_a = 0.4$ or develop symptoms with the remaining probability of $1-p_a = 0.6$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s=3$) and equal transmission rates.
Recovered individuals are assumed to be immune to reinfections throughout a semester.
Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}:
more specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
Assumed parameters are broadly consistent with other models of SARS-CoV-2 \citep{brett2020transmission,lavezzo2020suppression}.

Transmission processes are modeled by first setting the contact reproduction number $\Rc$, which we define as the average number of infectious contacts an infected individual would make throughout the course of their infection;
here, infectious contacts refer to contacts that would result in infection when the contactee (i.e., the contacted individual) is susceptible to infection.
The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
We further decompose $\Rc$ into pre-symptomatic $\mathcal R_p = \beta_p D_p$ and (a)symptomatic $\mathcal R_s = \beta_s D_s$ reproduction numbers, where $\beta_p$ and $\beta_s$ represent the corresponding infectious contact rates during pre-symptomatic and (a)symptomatic stages, respectively. Pre-symptomatic and (a)symptomatic reproduction numbers are calculated based on the assumed value of the proportion of presymptomatic transmission $p_p = 0.5$: $\mathcal R_p/\mathcal R_s = p_p/(1 - p_p)$.
On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events \citep{10.12688/wellcomeopenres.15842.3}.
Infectious contacts from local community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, they have a reduced probability of infection corresponding to their susceptibility (discussed later).

@@

In this study, we use this model to retrospectively analyze past outbreaks.
First, we try to match our model to epidemic patterns seen on campus for the first three semesters, during which there was limited campus transmission, by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume that all other parameters are equal across different groups for simplicity. 
We further assume that the population mixes homogeneously.
While these assumptions are most conservative, epidemiological parameters  and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.

@@

\section*{The spread of the Omicron variant on campus}

Epidemiological and interventions conditions changed throughout the spring semester of the 2021--2022 academic year.
We therefore extend to model to account for more realistic settings and focus on the outbreak patterns among undergraduate students.
First, based on \citep{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the Omicron variant by 10\% and 70\%, respectively. 
We also assume that the transmissibility of Omicron is reduced proportionally following previously assumed 90--to--20 ratio for the Delta variant;
in other words, two and three doses of  vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
The immunity from the third dose is assumed to take 7 days to develop and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
Finally, the isolation period is reduced to 5 days.
